Composable Generalized Consensus

### 2016-01-11 paper/cgc.tex
working on the writeup of the cgc proof.

Generalized Consensus~\cite{Lamport05GeneralizeConsensus} is a formal specification of the requirements that the distributed algorithm $A$ must meet in order for SMR to provide the illusion of a centralized fault-tolerant service.
Generalized Consensus uses the notion of a set of command-structures, or c-structs. C-structs are obtained by appending commands to the c-struct $\bot$. Given a c-struct $s$ and a command $c$, $s \bullet c$ is the c-struct obtained by appending $c$ to $s$. Given a sequence of commands $cs$, $s\star cs$ is the c-struct obtained by repeatedly appending the commands in $cs$ in the order where they appear in $cs$. By appropriately defining the operator $\bullet$ according to the
definition of $M$, C-structs can describe sequences of commands up the reordering of commuting
commands: it is possible to have $s \star \left[c_1,c_2\right] = s \star \left[c_2,c_1\right]$.
Moreover, c-structs must satisfy a number of properties: \setword{property 1}{prop:1}, c-structs must form a partial order under the operator $\preceq$ defined as: $s_1 \preceq s_2$ if and only if there exists a sequence of commands $cs$ such that $s_2 = s_1 \star cs$; \setword{property 2}{prop:2}, the greatest lower bound of a set of c-structs $S$ must always exists and be constructible from the same commands as the c-structs in $S$; and \setword{property 3}{prop:3}, a set of c-structs $S$ which has an upper bound, called a compatible set, must have a lowest upper bound which is constructible from the same commands as the c-structs in $S$.

@@

Generalized Consensus requires the following properties: non-triviality, stating that for any learner $l$, there is at all times a sequence of proposed commands $cs$ such that $learned\left[l\right] = \bot \bullet cs$; stability, stating that for any learner $l$, $learned\left[ l \right]$ increase with time; agreement, stating that $\left\{ learned\left[ l \right] :  l \in L\right\}$ is always compatible; and liveness, stating that for any proposed command $c$ and learner $l$, a c-struct of the form $s\bullet c$ is eventually learned by $l$.
Note that the FLP theorem~\cite{FischerLynchPaterson83ImpossibilityDistributedConsensusOneFaultyProcess} implies that the liveness property cannot be achieved in a networked system where processes may fail. Therefore, the liveness property guaranteed by Generalized Consensus algorithms is necessary weaker that the liveness property state above, maybe adding assumptions about faults and synchrony.
We refer the reader to the work of Lamport~\cite{Lamport05GeneralizeConsensus} for a thorough discussion of Generalized Consensus.

@@

Composable Generalized Consensus augments the Generalized Consensus specification with an abort interface that allows an SMR module to abort its execution, and with an init interface which allows an SMR module to start its execution where a preceding aborted module left off.
To specify the composition interface, we introduced a third type of processes called the \emph{switchers}. Switchers can receive \emph{init values} from a preceding aborting modules and can pass \emph{abort values} to pass the baton to a next SMR module when the current SMR module aborts.

We describe the behavior of switchers using the variable $from$, which contains all the init values received by any switcher, and the variable $to$, which contains all the abort values produced by the switchers.
The variable $prop$ is used as before, but the variable $learned$ now maps learners to sets of c-structs and, for each learner $l$, $learned\left[ l \right]$ consists of all the c-structs learned so far by $l$.
We define the set of valid c-structs as the c-structs of the form $t\star cs$ where $cs$ is a sequence of proposed commands, and $t$ is the greatest lower bound of a non-empty subset of the init values received.

Composable Generalized Consensus requires that no value be learned unless at least one init value have been received (Initialization); that for any learner $l$, $learned\left[ l \right]$ is valid (Nontriviality); that $\left\{ learned\left[ l \right] :  l \in L\right\}$ is always compatible (Agreement); that every abort value is valid, that for any abort value $s_a$ and learned value $s_l$ we have $s_l \preceq s_a$ (Safe Abort); and that for any proposed command $c$ and learner $l$, a c-struct
of the form $s\bullet c$ is eventually learned by $l$ or a switcher eventually produces an abort value (Liveness).
A TLA$^+$ specification of CGC appears in \cref{fig:cgc}.

We say that an algorithm implementing CGC under some refinement mapping is a \emph{CGC instance}, and that two CGC instances $i_1$ and $i_2$ are compatible if they update disjoint variables except for the $to$ variable of $i_1$ and the $from$ variable of $i_2$, which must be identical.
Note that in the execution of two compatible instances, any abort value produced by the first instance is an init value for the second instance.
We say that a CGC instance is an initial instance when its set of init values is always a subset of $\left\{ \bot \right\}$.

@@

First note that the following invariants hold of the execution of $Spec$: invariant 1, stating that any learned c-struct of $CGC1$ is smaller than any abort c-struct of $CGC1$; invariant 2, stating that for any non-empty subset $S$ of the abort c-structs of $CGC1$, there is a sequence $cs$ of commands proposed in $CGC1$ such that $\operatorname{GLB}\left( S \right) = \bot\star cs$; and invariant 3, stating that for every learned c-struct or abort c-struct $s$ of $CGC2$, there is a non-empty subset $S$ of the init c-structs of $CGC2$ and a sequence of commands proposed in $CGC2$ such that $s  = \operatorname{GLB}\left( S \right)\star cs$.
Invariant 1 is enforced by the first conjunct of the $Learn(l)$ transition of $CGC1$, and invariant 3 is enforced by the quantifier bound of the $Learn(l)$ and $Abort$ transition of $CGC2$, using the definition of $Valid$. 
To see why invariant 2 holds, note that the quantifier bound of the $Abort$ transition of $CGC1$, the fact that the only init c-struct of $CGC1$ is $\left\{ \bot \right\}$, and the definition of $Valid$ ensure that any abort c-struct of $CGC1$ is of the form $\bot\star cs$ where $cs$ is a sequence of proposed commands of $CGC1$. Then, we obtain invariant 2 using the~\ref{prop:2} of c-structs.

We now prove that each transition of $Spec$ is a transition of $CGC!Spec$:
The transition $CGC1!Propose$ inserts a command $c$ in the set $CCG1!prop$, and, since $CGC!prop = CGC1!prop \cup CGC2!prop$, the transition inserts $c$ in the set $CGC!prop$; therefore, $CGC1!Propose$ implies $CGC!Propose$.
By an analogous reasoning, $CGC2!Propose$ implies $CGC!Propose$.
The transition $CGC1!InitM$ inserts $\bot$ in the set $CGC1!from$ (because $CGC1$ is an initial instance), and, since $CGC!prop = CGC1!prop$, the transition also inserts $\bot$ in $CGC!prop$; therefore, $CGC1!InitM$ implies $CGC!InitM$.
The $CGC1!Abort$ transition does not modify any variable of the module instance $CGC$; therefore, $CGC1!Abort$ is a stuttering transition of $CGC$.
Similarly, the $CGC2!InitM$ transition does not modify any variable of the module instance $CGC$; therefore, $CGC2!InitM$ is a stuttering transition of $CGC$.
TODO\@: complete.
We have formalized informal proof above in Isabelle/HOL (and not TLAPS~\cite{CousineauETAL12TlaProofs}, because the authors are proficient in Isabelle/HOL but not in TLAPS); the proof is available at \url{http://losa.fr/cgc}.

>>>>

Generalized Consensus~\cite{Lamport05GeneralizeConsensus} is a formal specification of the requirements that the distributed algorithm $A$ must meet in order for SMR to provide the illusion of a centralized fault-tolerant service.
Generalized Consensus uses the notion of a set of command-structures, or c-structs. C-structs are obtained by appending commands to the c-struct $\bot$. Given a c-struct $s$ and a command $c$, $s \bullet c$ is the c-struct obtained by appending $c$ to $s$. Given a sequence of commands $cs$, $s\star cs$ is the c-struct obtained by repeatedly appending the commands in $cs$ in the order where they appear in $cs$. By appropriately defining the operator $\bullet$ according to the
definition of $M$, C-structs can describe sequences of commands up the reordering of commuting
commands: it is possible to have $s \star \left[c_1,c_2\right] = s \star \left[c_2,c_1\right]$.
Moreover, c-structs must satisfy a number of properties: \setword{property 1}{prop:1}, c-structs must form a partial order under the operator $\preceq$ defined as: $s_1 \preceq s_2$ if and only if there exists a sequence of commands $cs$ such that $s_2 = s_1 \star cs$; \setword{property 2}{prop:2}, the greatest lower bound of a set of c-structs $S$ must always exists and be constructible from the same commands as the c-structs in $S$; and \setword{property 3}{prop:3}, a set of c-structs $S$
which has an upper bound, called a \emph{compatible} set, must have a lowest upper bound which is constructible from the same commands as the c-structs in $S$.

@@

Generalized Consensus requires the following properties: \emph{non-triviality}, stating that for any learner $l$, there is at all times a sequence of proposed commands $cs$ such that $learned\left[l\right] = \bot \bullet cs$; \emph{stability}, stating that for any learner $l$, $learned\left[ l \right]$ increase with time; \emph{agreement}, stating that $\left\{ learned\left[ l \right] :  l \in L\right\}$ is always \emph{compatible}; and \emph{liveness}, stating that for any proposed command $c$ and learner $l$, a c-struct of the form $s\bullet c$ is eventually learned by $l$.
Note that the FLP theorem~\cite{FischerLynchPaterson83ImpossibilityDistributedConsensusOneFaultyProcess} implies that the liveness property cannot be achieved in a networked system where processes may fail. Therefore, the liveness property guaranteed by Generalized Consensus algorithms is necessary weaker that the liveness property state above, maybe adding assumptions about faults and synchrony.
We refer the reader to the work of Lamport~\cite{Lamport05GeneralizeConsensus} for a thorough discussion of Generalized Consensus.

@@

Composable Generalized Consensus augments the Generalized Consensus specification with an abort interface that allows an SMR module to abort its execution, and with an init interface which allows an SMR module to start its execution where a preceding aborted module left off.
To specify the composition interface, we introduced a third type of processes called the \emph{switchers}. Switchers can receive \emph{init c-structs} from a preceding aborting modules and can pass \emph{abort c-structs} to pass the baton to a next SMR module when the current SMR module aborts.

We describe the behavior of switchers using the variable $from$, which contains all the init c-structs received by any switcher, and the variable $to$, which contains all the abort c-structs produced by the switchers.
The variable $prop$ is used as before, but the variable $learned$ now maps learners to sets of c-structs and, for each learner $l$, $learned\left[ l \right]$ consists of all the c-structs learned so far by $l$.
We define the set of valid c-structs as the c-structs of the form $t\star cs$ where $cs$ is a sequence of proposed commands, and $t$ is the greatest lower bound of a non-empty subset of the init c-structs received.

Composable Generalized Consensus requires the following properties: \emph{initialization}, stating that no value be learned unless at least one init c-struct have been received; \emph{non-triviality}, stating that for any learner $l$, $learned\left[ l \right]$ is valid; \emph{agreement}, stating that $\left\{ learned\left[ l \right] :  l \in L\right\}$ is always compatible; \emph{safe-abort}, stating that every abort c-struct is valid, that for any abort c-struct $s_a$ and learned value $s_l$ we have
$s_l \preceq s_a$; and \emph{liveness}, stating that for any proposed command $c$ and learner $l$, a c-struct of the form $s\bullet c$ is eventually learned by $l$ or a switcher eventually produces an abort c-struct.
A TLA$^+$ specification of the safety properties of CGC appears in \cref{fig:cgc}.

We say that an algorithm implementing CGC under some refinement mapping is a \emph{CGC instance}, and that two CGC instances $i_1$ and $i_2$ are compatible if they update disjoint variables except for the $to$ variable of $i_1$ and the $from$ variable of $i_2$, which must be identical.
Note that in the execution of two compatible instances, any abort c-struct produced by the first instance is an init c-struct for the second instance.
We say that a CGC instance is an initial instance when its set of init c-structs is always a subset of $\left\{ \bot \right\}$.

@@

First note that the following invariants hold of the execution of $Spec$: \emph{invariant 1}, stating that any learned c-struct of $CGC1$ is smaller than any abort c-struct of $CGC1$; \emph{invariant 2}, stating that for any non-empty subset $S$ of the abort c-structs of $CGC1$, there is a sequence $cs$ of commands proposed in $CGC1$ such that $\operatorname{GLB}\left( S \right) = \bot\star cs$; and \emph{invariant 3}, stating that for every learned c-struct or abort c-struct $s$ of $CGC2$, there is a non-empty subset $S$ of the init c-structs of $CGC2$ and a sequence of commands proposed in $CGC2$ such that $s  = \operatorname{GLB}\left( S \right)\star cs$.
Invariant 1 is enforced by the first conjunct of the $Learn(l)$ transition of $CGC1$, and invariant 3 is enforced by the quantifier bound of the $Learn(l)$ and $Abort$ transition of $CGC2$, using the definition of $Valid$. 
To see why invariant 2 holds, note that the quantifier bound of the $Abort$ transition of $CGC1$, the fact that the only init c-struct of $CGC1$ is $\left\{ \bot \right\}$, and the definition of $Valid$ ensure that any abort c-struct of $CGC1$ is of the form $\bot\star cs$ where $cs$ is a sequence of proposed commands of $CGC1$. Then, we obtain invariant 2 using the~\ref{prop:2} of c-structs.

We now prove that each transition of $Spec$ is a transition (i.e.\ implies a transition) of $CGC!Spec$, analysing the transition one by one.
Note that the transition $CGC1!Propose$ inserts $c$ in the set $CGC!prop$, because $CGC!prop = CGC1!prop \cup CGC2!prop$; therefore, $CGC1!Propose$ implies $CGC!Propose$.
Similarly, the $CGC2!Propose$ implies $CGC!Propose$.
Note that the transition $CGC1!InitM$ inserts $\bot$ in the set $CGC!from$, because $CGC1$ is an initial instance and $CGC!from = CGC1!from$; therefore, $CGC1!InitM$ implies $CGC!InitM$.
The $CGC1!Abort$ transition does not modify any variable of the module instance $CGC$; therefore, $CGC1!Abort$ is a stuttering transition of $CGC$.
Similarly, the $CGC2!InitM$ transition is a stuttering transition of $CGC$.

Note that the $CGC1!Learn(l)$ transition inserts a c-struct $s$ in the set $CGC!learned\left[ l \right]$, because $CGC!leanred[l] = CGC1!learned\left[ l \right] \cup CGC2!learned\left[ l \right]$; therefore, to show that $CGC1!Learn\left( l \right)$ implies $CGC!Learn\left( l \right)$, we must show (1), that $s$ is smaller than any abort c-struct of $CGC$, (2), that $s$ is compatible with any learned c-struct of $CGC$, and (3), that $s$ is valid in $CGC$. 
To show (1), consider an abort c-struct $a$ of $CGC2$. By invariant 3, $a$ is of the form $\operatorname{GLB}\left( S \right)\star cs$ where $S$ is a subset of the init c-structs of $CGC2$. Therefore, $\operatorname{GLB}\left( S \right)\preceq a$. Moreover, $s$ is smaller than any init c-struct of $CGC2$, by invariant 1. Therefore, $\operatorname{GLB}\left( S \right)$ is bigger than $s$. By transitivity of $\preceq$, we get that $s$ is smaller than $a$.
To show (2), consider a learned c-struct $s_l$ of $CGC$. There are two cases: (2a), $s_l$ is a learned c-struct of $CGC1$, and (2b), $s_l$ is a learned c-struct of $CGC2$. In case (2a), the second conjunct of the $CGC1!Learn(l)$ transition establishes that $s$ and $s_l$ are compatible. In case (2b), by reasoning as in (1), we establish that $s \preceq s_l$. Therefore, $s$ and $s_l$ are compatible.
To show (3), observe that any c-struct valid in $CGC1$ is also valid in $CGC2$.

Similarly, the $CGC2!Learn(l)$ transition inserts a valid c-struct $s$ in the set $CGC!learned\left[ l \right]$, and we must show the same 3 properties.
Property (1) is established by the first conjunct of the $CGC2!Learn\left( l \right)$ transition, using the fact that $CGC!to = CGC2!to$.
To show (2), consider a learned c-struct of $CGC$.
There are two cases: (2a), $s_l$ is a learned c-struct of $CGC1$, and (2b), $s_l$ is a learned c-struct of $CGC2$. In case (2b), the second conjunct of the $CGC2!Learn(l)$ transition establishes that $s$ and $s_l$ are compatible. In case (2a), by reasoning similarly as in TODO use an invariant...
To show (3)\ldots
TODO\@: complete.
We have formalized informal proof above in Isabelle/HOL (and not TLAPS~\cite{CousineauETAL12TlaProofs}, because the authors are proficient in Isabelle/HOL but not in TLAPS); the proof is available at \url{http://losa.fr/cgc}.

### 2016-01-11 paper/intro.tex
working on the writeup of the cgc proof.

An SMR module is an SMR algorithm augmented with a \emph{composition interface}
and satisfying our correctness condition, called Composable Generalized
Consensus (CGC). The composition interface consists of an \emph{abort interface}
for aborting an SMR module's execution and passing the baton to another
SMR module and with a matching \emph{init interface} for taking over the
execution of an aborting SMR module. Two SMR modules are composed by passing
the outputs produced by the abort interface of the first SMR module, called
abort values, as inputs to the init interface of the second SMR module,
called init values in the context of the second SMR module. A SMR module
must satisfy a correctness condition that we call Composable Generalized
Consensus (abbreviated CGC). Any correct SMR module is a safe Generalized
Consensus~\cite{Lamport05GeneralizeConsensus} algorithm (i.e.\ a kind of SMR
algorithm), but the progress guarantees may vary between SMR modules: CGC only
requires a SMR module to make progress executing commands or to abort its
execution. CGC also relates the init values, learned values (learned values
represent the local state-machine state of individual replicas), and abort
values in order for our composition theorem to hold.
%Note that Generalized Consensus is a specification of SMR algorithms that takes into account the commutativity relation among commands and can offer important performance advantages in practice.

@@

In practice, we provide a TLA+ specification of CGC that the user of our framework can use to debug and prove new modules correct using the TLA$^+$ toolbox~\cite{TLATools}.
Moreover, we also provide two TLA+ specifications simplifying the development of modules based on Refined Quorum Systems~\cite{GuerraouiVukolic10RefinedQuorumSystems}. These two specifications both implement CGC and provide a skeleton for a user to build her own module. The two specifications demonstrate how to update the local state of a replica after gathering the states or the init values of a write quorum of replicas, and how to
produce an abort value after gathering the state of a read quorum of replicas. Each of the specifications achieves a particular consistency/latency trade-off~\cite{Abadi12ConsistencyTradeoffsModernDistributedDatabaseSystem}.
The two specifications use different combinations of sizes of read and write quorums, which affects the invariants that the replicas must maintain and, in consequence, the latency with which a command can be processed but the module.
Our two specifications abstract over how to disseminate the commands to the replicas (e.g.\ whether to use a leader as in Paxos, multiple leaders as in MultiCoordinatedPaxos, or whether to broadcast commands without using a leader as in Fast Paxos).
The user can therefore specify a concrete command dissemination procedure and then model-check that the obtained algorithm refines one of the two specifications.  
We demonstrate the practical use of our framework by presenting modules based on Paxos~\cite{Lamport98ParttimeParliament}, Fast Paxos\cite{Lamport06FastPaxos}, Chain Replication~\cite{RenesseSchneider04ChainReplicationSupportingHighThroughputAvailability}, a case of Vertical Paxos~\cite{LamportMalkhiZhou09VerticalPaxosPrimarybackupReplication}, and EPaxos~\cite{MoraruAndersenKaminsky13ThereIsMoreConsensusEgalitarianParliaments}
(TODO, but would be nice).

>>>>

An SMR module is an SMR algorithm augmented with a \emph{composition interface}
and satisfying our correctness condition, called Composable Generalized
Consensus (CGC). The composition interface consists of an \emph{abort interface}
for aborting an SMR module's execution and passing the baton to another
SMR module and with a matching \emph{init interface} for taking over the
execution of an aborting SMR module. Two SMR modules are composed by passing
the outputs produced by the abort interface of the first SMR module, called
abort c-structs, as inputs to the init interface of the second SMR module,
called init c-structs in the context of the second SMR module. A SMR module
must satisfy a correctness condition that we call Composable Generalized
Consensus (abbreviated CGC). Any correct SMR module is a safe Generalized
Consensus~\cite{Lamport05GeneralizeConsensus} algorithm (i.e.\ a kind of SMR
algorithm), but the progress guarantees may vary between SMR modules: CGC only
requires a SMR module to make progress executing commands or to abort its
execution. CGC also relates the init c-structs, learned values (learned values
represent the local state-machine state of individual replicas), and abort
values in order for our composition theorem to hold.
%Note that Generalized Consensus is a specification of SMR algorithms that takes into account the commutativity relation among commands and can offer important performance advantages in practice.

@@

In practice, we provide a TLA+ specification of CGC that the user of our framework can use to debug and prove new modules correct using the TLA$^+$ toolbox~\cite{TLATools}.
Moreover, we also provide two TLA+ specifications simplifying the development of modules based on Refined Quorum Systems~\cite{GuerraouiVukolic10RefinedQuorumSystems}. These two specifications both implement CGC and provide a skeleton for a user to build her own module. The two specifications demonstrate how to update the local state of a replica after gathering the states or the init c-structs of a write quorum of replicas, and how to
produce an abort c-struct after gathering the state of a read quorum of replicas. Each of the specifications achieves a particular consistency/latency trade-off~\cite{Abadi12ConsistencyTradeoffsModernDistributedDatabaseSystem}.
The two specifications use different combinations of sizes of read and write quorums, which affects the invariants that the replicas must maintain and, in consequence, the latency with which a command can be processed but the module.
Our two specifications abstract over how to disseminate the commands to the replicas (e.g.\ whether to use a leader as in Paxos, multiple leaders as in MultiCoordinatedPaxos, or whether to broadcast commands without using a leader as in Fast Paxos).
The user can therefore specify a concrete command dissemination procedure and then model-check that the obtained algorithm refines one of the two specifications.  
We demonstrate the practical use of our framework by presenting modules based on Paxos~\cite{Lamport98ParttimeParliament}, Fast Paxos\cite{Lamport06FastPaxos}, Chain Replication~\cite{RenesseSchneider04ChainReplicationSupportingHighThroughputAvailability}, a case of Vertical Paxos~\cite{LamportMalkhiZhou09VerticalPaxosPrimarybackupReplication}, and EPaxos~\cite{MoraruAndersenKaminsky13ThereIsMoreConsensusEgalitarianParliaments}
(TODO, but would be nice).
